A New Metropolis-Hastings Test for Fast Minibatch MCMC Inference and Optimization

### 2016-05-16 nips_2016.tex
add temptuerare

In this section, we introduce our new Metropolis-Hastings test and show how to use it to achieve
sufficient mixing and/or to reach better posterior modes. To start, we introduce two key values we
will use frequently: $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta')\prod_{i=1}^n p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^n p(x_i \mid \theta_t)q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}, so the terms here
consider the likelihood ratios in log space. The only difference between $\Delta$ and $\Delta'$ is
that they use $N$ and $n$ terms in the likelihood ratio, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
If we fix $\theta_t$ and $\theta'$, the randomness in our $\Delta'$ samples only comes from the
minibatch. Since the minibatch is chosen randomly, the distribution of $\sum_{i=1}^n (\log p(x_i\mid
\theta') - \log p(x_i\mid \theta_t))$ converge to a Gaussian by the Central Limit Theorem.
\end{proof}

@@

\begin{lemma}
For one step sampling from $\theta_t$ to $\theta'$, if the jumping step  $|\theta_t - \theta'| <
\epsilon$, and the gradient of the log-likelihood is bounded by a constant factor, i.e. $|\nabla
(\log \Pr(x_i| \theta))| < k$, the variance of $\Delta'$ is bounded by $4\epsilon^2 k^2 m$, where
$m$ is the minibatch size.
\end{lemma}

\begin{proof}
From Lemma~\ref{lem:gaussian}, only the $\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid
\theta_t))$ term brings randomness into $\Delta'$. To simplify the subsequent notation, we define
$g_i(\theta) = \log p(x_i\mid \theta)$. Then by Taylor's Theorem:
\[
|g(\theta') - g(\theta_t)| = |\nabla g(\theta_t) (\theta' - \theta_t) + o(\theta' - \theta_t)^2| \leq 2\epsilon k.
\]
Thus, with given the $\theta$, and $\theta'$, the variance of $\Delta'$ can be write as:
\[
{\rm Var}(\Delta') = {\rm Var}\left[\sum_{i=1}^n g_i(\theta') - g_i(\theta_t)\right] \leq
\sum_{i=1}^n {\rm Var}(g_i(\theta') - g_i(\theta_t)) \leq  4\epsilon^2 m k^2,
\]
as desired.
\end{proof}

Since ${\rm Var}(\Delta') < \epsilon^2 k^2 m$, we can assume the maximum value of $X_{\xi}$ is
proportional to the standard deviation of $\Delta$, i.e. $\epsilon$. Thus, we can assume
$\zeta=\epsilon C$, where $C$ is a constant factor.

\begin{theorem}
Assume the stationary distribution $\pi$ satisfies $\|T \circ P_0 - \pi\| \leq \eta \|P_0 - \pi\|$,
where $\eta \in [0, 1)$ for all distribution $P_0$. And assume there exists $\rho_t < 1$ such that
$\|T'_i(x, \cdot) - \pi_i'\| \leq 2\rho_i$. We can get $\| T_t' \circ T_{t-1}' \circ \cdots \circ T_1' P_0
- \pi_0 \| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s \alpha_s + \alpha_t $,
where $\alpha_i = \frac{\epsilon_i k }{1-\eta}$.
\end{theorem}

\begin{proof}
First, by Theorem 1 in~\cite{cutting_mh_2014}, $\|T \circ P_0 - \pi\| \leq \eta \|P_0
- \pi\|$ and $\|T-T'_i\| \leq \epsilon_i k$, the norm between the stationary distribution is $\|\pi
- \pi_i'\|\leq \frac{\epsilon_i k}{1-\eta}$.

Second, by Theorem 3.6 in~\cite{yang2013sequential}, we have  $\| T_t' \circ T_{t-1}' \circ \cdots
\circ T_1' P_0 - \pi_t \| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s
\alpha_s$, where $\alpha = \frac{\epsilon_i k}{1-\eta}$. Thus, we get:
\begin{align*}
 \| T_t' \circ \cdots \circ T_1' P_0 - \pi_0 \| &\leq \|T_t' \circ \cdots \circ T_1' P_0 - \pi_t\| + \|\pi_t - \pi_0\| \\
 &\leq \sum_{s=1}^t \left\{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\right\} \rho_s \alpha_s + \alpha_t,
\end{align*}
as desired.
\end{proof}

>>>>

In this section, we introduce our new Metropolis-Hastings test and show how to use it to achieve
sufficient mixing and/or to reach better posterior modes. To start, we introduce two key values we
will use frequently: $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta') \left ( \prod_{i=1}^N p(x_i \mid \theta') \right)^t q(\theta_t \mid
\theta')}{p(\theta_t)\left (\prod_{i=1}^N p(x_i \mid \theta_t) \right) ^t q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') \left ( \prod_{i=1}^n p(x_i \mid \theta') \right ) ^t q(\theta_t \mid
\theta')}{p(\theta_t) \left ( \prod_{i=1}^n p(x_i \mid \theta_t) \right ) ^t q(\theta' \mid
\theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}, so the terms here
consider the likelihood ratios in log space. The only difference between $\Delta$ and $\Delta'$ is
that they use $N$ and $n$ terms in the likelihood ratio, respectively. We assume $n \ll N$ (so
$\Delta'$ is substantially faster to compute) and that the $n$ data points $x_i$ represent a random
minibatch of data.

@@

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\frac{1}{t}\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
If we fix $\theta_t$ and $\theta'$, the randomness in our $\Delta'$ samples only comes from the
minibatch. Since the minibatch is chosen randomly, the distribution of $\frac{1}{t}\sum_{i=1}^n (\log p(x_i\mid
\theta') - \log p(x_i\mid \theta_t))$ converge to a Gaussian by the Central Limit Theorem.
\end{proof}

@@

\begin{lemma}
For one step sampling from $\theta_t$ to $\theta'$, if the jumping step  $|\theta_t - \theta'| <
\epsilon$, and the gradient of the log-likelihood is bounded by a constant factor, i.e. $|\nabla
(\log \Pr(x_i| \theta))| < k$, the variance of $\Delta'$ is bounded by $4\epsilon^2 k^2 m /t^2$, where
$m$ is the minibatch size, and $t$ is the temperature.
\end{lemma}

\begin{proof}
From Lemma~\ref{lem:gaussian}, only the $\frac{1}{t}\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid
\theta_t))$ term brings randomness into $\Delta'$. To simplify the subsequent notation, we define
$g_i(\theta) = \log p(x_i\mid \theta)$. Then by Taylor's Theorem:
\[
|g(\theta') - g(\theta_t)| = |\nabla g(\theta_t) (\theta' - \theta_t) + o(\theta' - \theta_t)^2| \leq 2\epsilon k.
\]
Thus, with given the $\theta$, and $\theta'$, the variance of $\Delta'$ can be write as:
\[
{\rm Var}(\Delta') = {\rm Var}\left[\frac{1}{t}\sum_{i=1}^n g_i(\theta') - g_i(\theta_t)\right] \leq
\sum_{i=1}^n \frac{1}{t^2} {\rm Var}(g_i(\theta') - g_i(\theta_t)) \leq  4\epsilon^2 m k^2 /t^2,
\]
as desired.
\end{proof}

Since ${\rm Var}(\Delta') < \epsilon^2 k^2 m / t^2$, we can assume the maximum value of $X_{\xi}$ is
proportional to the standard deviation of $\Delta$, i.e. $\epsilon / t$. Thus, we can assume
$\zeta=\epsilon C/t$, where $C$ is a constant factor.

\begin{theorem}
Assume the stationary distribution $\pi$ satisfies $\|T \circ P_0 - \pi\| \leq \eta \|P_0 - \pi\|$,
where $\eta \in [0, 1)$ for all distribution $P_0$. And assume there exists $\rho_t < 1$ such that
$\|T'_i(x, \cdot) - \pi_i'\| \leq 2\rho_i$. We can get $\| T_t' \circ T_{t-1}' \circ \cdots \circ T_1' P_0
- \pi_0 \| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s \alpha_s + \alpha_t $,
where $\alpha_i = \frac{\epsilon_i C }{t_i(1-\eta)}$.
\end{theorem}

\begin{proof}
First, by Theorem 1 in~\cite{cutting_mh_2014}, $\|T \circ P_0 - \pi\| \leq \eta \|P_0
- \pi\|$ and $\|T-T'_i\| \leq \epsilon_i C/t_i$, the norm between the stationary distribution is $\|\pi
- \pi_i'\|\leq \frac{\epsilon_i C}{t_i(1-\eta)}$.

Second, by Theorem 3.6 in~\cite{yang2013sequential}, we have  $\| T_t' \circ T_{t-1}' \circ \cdots
\circ T_1' P_0 - \pi_t \| \leq \sum_{s=1}^t \{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\} \rho_s
\alpha_s$, where $\alpha = \frac{\epsilon_i C}{t_i(1-\eta)}$. Thus, we get:
\begin{align*}
 \| T_t' \circ \cdots \circ T_1' P_0 - \pi_0 \| &\leq \|T_t' \circ \cdots \circ T_1' P_0 - \pi_t\| + \|\pi_t - \pi_0\| \\
 &\leq \sum_{s=1}^t \left\{\prod _{u=s+1}^t \rho_u (1-\alpha_u)\right\} \rho_s \alpha_s + \alpha_t,
\end{align*}
as desired.
\end{proof}

The Theorem above indicates that the difference of the distburition obtained by applying the
